\morus: Biases in the Stream

### 2018-05-10 morusAC_00_Abstract.tex
homogenized references
minor

\begin{abstract}
  \morus is a high-performance authenticated encryption algorithm
  submitted to the CAESAR competition, and recently selected as a
  finalist.  There are three version of \morus: \morus[640] with a
  640-bit state and a 128-bit key, and \morus[1280] with a 1280-bit
  state and a 128-bit or 256-bit key; for all versions the security
  claim for confidentiality matches the key size.  In this paper, we
  analyze the components of this algorithm (initialization, state
  update and tag generation), and report several results.

@@

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify the bias, we exploit rotational symmetries in \morus
  using linear masks that are invariants by word-rotations of the
  state.  This motivates us to introduce word-oriented versions of
  \morus called \minimorus, which help the
  analysis. The attack has been implemented and verified on
  \minimorus, having a bias of $2^{-16}$.

@@

  We also study reduced versions of the initialization and
  finalization of \morus to evaluate the security margin of these
  components.  We show a forgery attack when the finalization is
  reduced from 10 rounds to 3, and a key-recovery attack in the
  nonce-misue setting when the initialization is reduced from 16 to 10
  steps.  These additional results do not threaten the full \morus but
  studying all aspects of the design is useful to understand its
  strengths and weaknesses.

>>>>

%!TEX root = morusAC.tex

\begin{abstract}
  \morus is a high-performance authenticated encryption algorithm
  submitted to the CAESAR competition, and recently selected as a
  finalist.  There are three versions of \morus: \morus[640] with a
  640-bit state and a 128-bit key, and \morus[1280] with a 1280-bit
  state and a 128-bit or 256-bit key; for all versions the security
  claim for confidentiality matches the key size.  In this paper, we
  analyze the components of this algorithm (initialization, state
  update and tag generation), and report several results.

@@

% More precisely, we show that a linear correlation exists between some bits of the plaintext and ciphertext. Moreover the bias depends purely on the plaintext, and not on the secret key nor the nonce, hence restricting the single use of nonce or even rekeying does not prevent the attacks.
%
% From the nature of the linear cryptanalysis, the attack works in the known plaintext attacks. Moreover, if a fixed plaintext is encrypted multiple times, the correlation can be used to recover a 1-bit information of the plaintext. The attack principally has larger impact because it can be used to recover more unknown bits of the plaintext, provided an initial segment of the plaintext is known. 
%
  To identify the bias, we exploit rotational symmetries in \morus
  using linear masks that are invariant by word-rotations of the
  state.  This motivates us to introduce word-oriented versions of
  \morus called \minimorus, which help the
  analysis. The attack has been implemented and verified on
  \minimorus, where it yields a bias of $2^{-16}$.

@@

  We also study reduced versions of the initialization and
  finalization of \morus to evaluate the security margin of these
  components.  We show a forgery attack when the finalization is
  reduced from 10 rounds to 3, and a key-recovery attack in the
  nonce-misue setting when the initialization is reduced from 16 to 10
  steps.  These additional results do not threaten full \morus, but
  studying all aspects of the design is useful to understand its
  strengths and weaknesses.
